---
title: "Define Episodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Define Episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

When longitudinal health records are collected over months or years, distinct treatment episodes are commonly identified based on periods of inactivity between observations. This involves specifying a time interval long enough to sufficiently distinguish the end of one treatment episode from the beginning of another.

`leap` supports this time-based approach to delimiting treatment episodes while providing greater flexibility than simply specifying a time delimiter a priori. Specifically, leap provides tools for examining the empirical distribution of time between sessions and estimating data-informed episode delimiters. This article describes these approaches and demonstrates how alternative delimiters can be evaluated and compared.

# Predefined delimiter

The simplest approach is to specify an a priori amount of time between consecutive sessions that constitutes the beginning of a new treatment episode. By default, leap uses a 90-day delimiter. When the time since a client's preceding session meets or exceeds this threshold, the subsequent session is assigned to a new episode.

The default can be applied by first calculating the number of days between consecutive sessions using add_session_lag() and then assigning episode membership with add_episode_id():

```{r, default lag, eval = FALSE}
# by default, episodes are distinguished 
# by 90 days of inactivity
episode_data <- data |>
  add_session_lag() |>
  add_episode_id()
```

Researchers may instead specify a different delimiter when another threshold is theoretically or clinically appropriate, or when examining the sensitivity of episode definitions to alternative thresholds. For example, a 180-day delimiter can be applied as follows:

```{r bigger lag, eval = FALSE}
episode_data <- data |>
  add_session_lag() |>
  add_episode_id(delimiter = 180)
```

The choice of delimiter determines which gaps in treatment are interpreted as breaks between episodes and can therefore affect the number, duration, and composition of episodes identified in the data. When no established threshold is available, or when the appropriateness of a predefined threshold is uncertain, the observed distribution of session lags can also be used to inform this decision.

# Empirical delimiter

Rather than specifying an episode delimiter a priori, candidate thresholds can be estimated from the observed distribution of session lags. `lag_delimiter()` provides several approaches for doing so, including thresholds based on the standard deviation, interquartile range, and empirical quantiles of the observed session gaps.

```{r delimiter, eval = FALSE}
# standard deviation 
lag_delimiter(data, method = "sd")
# interquartile range
lag_delimiter(data, method = "iqr")
# quantile 
lag_delimiter(data, method = "quantile")
```

These approaches provide simple distribution-based thresholds that can be used to examine whether the observed spacing of sessions supports alternative definitions of treatment episodes.

# Estimated delimiter

`leap` also provides a mixture-modeling approach for estimating an episode delimiter:

```{r, estimated delimiters, eval = FALSE}
# mixture modeling 
lag_delimiter(data, method = "mixture")
```

The mixture-modeling approach, implemented using the `mclust` package [@scrucca2016], treats the distribution of session lags as arising from distinct underlying components. In the context of treatment episodes, these components are used to distinguish relatively short gaps that are more consistent with continued treatment from longer gaps that may indicate a new episode of care.

The estimated boundary between these components provides a data-informed candidate delimiter. Importantly, this value should not be interpreted as a definitive clinical boundary. Rather, it provides an empirical basis for evaluating how treatment episodes might be defined in a particular data set and can be considered alongside theoretical, clinical, or previously established thresholds.
