---
title: "Defining Treatment Episodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining Treatment Episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Episode identification

Treatment episodes are commonly defined using a predetermined period of inactivity, such as 90 days without treatment. `leap` supports this approach while also providing tools for examining the empirical distribution of time between sessions.

Session lags can be calculated and inspected directly:

```{r lags, eval = FALSE}
episode_data <- data |>
  add_session_lag() |>
  add_episode_id(delimiter = 90)
```

Candidate episode delimiters can also be estimated from the observed distribution of session lags:

```{r delimiter, eval = FALSE}
lag_delimiter(data, method = "iqr")
lag_delimiter(data, method = "quantile")
lag_delimiter(data, method = "mixture")
```

The mixture-modeling approach, implemented using the `mclust` package [@scrucca2016], treats episode identification as an empirical classification problem by identifying relatively short- and long-gap components in the observed distribution of session lags. Estimated delimiters should be interpreted as data-informed candidate thresholds rather than definitive clinical boundaries between treatment episodes.
