---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical `leap` workflow begins with longitudinal treatment data in which each row represents a treatment session for a client. From these records, `leap` provides tools for identifying distinct episodes of care, describing and visualizing the resulting episode structure, and analyzing change within and across treatment episodes.

This guide provides a brief introduction to the core `leap` workflow. More detailed discussions of episode identification, visualization, and statistical modeling are available in the package articles.

# Step 1: Prepare and check data

To get started, `leap` requires person-period longitudinal data, with one row representing each treatment session. At minimum, the data should contain variables identifying the client, session, session date, and outcome of interest.

The example below uses a small data set included with leap:

```{r setup, include = FALSE}

# packages
library(tidyverse)
devtools::load_all()

# load example data
load("~/Library/CloudStorage/Dropbox/01_desk/04_personal/02_data_science/02_projects/01_pkgs/leap/data/simulation_data.RData")

# select subset for vignette 
raw_df <- data %>%
  purrr::pluck("unbalanced", "clinical", "diminishing_resp") %>%
  dplyr::nest_by(client_id) %>%
  dplyr::ungroup() %>%
  dplyr::slice_head(n = 10) %>%
  tidyr::unnest(cols = data)

```

```{r}
head(raw_df)
```

`leap` uses a standardized set of variable names throughout its functions. If the original data use different names, `cols_standard()` can be used to rename the core variables before proceeding:

```{r, variables, eval = FALSE}
raw_df <- cols_standard(
  data = raw_df, 
  client = "client_id", 
  session = "session_id", 
  date = "session_date",
  outcome = "outcome"
  )
```

Once the variables have been standardized, `check_raw()` can be used to verify that the data contain the information required to identify treatment episodes:

```{r, check, eval = FALSE}
check_raw(raw_df)
```

The output from this initial check helps identify missing or improperly structured variables before episode-level variables are derived.

# Step 2: Identify treatment episodes

Once the session-level data have been prepared, leap can be used to identify distinct treatment episodes and derive the variables needed for subsequent episode analyses. The `check_` functions evaluate the structure of the data before and after episode identification, while the `add_` functions derive treatment episode variables.

The necessary episode variables can be added individually or created together using the wrapper `add_episode_vars()`. The complete workflow is shown below:

```{r identify}
# Identify treatment episodes and derive episode variables
episode_df <- add_episode_vars(raw_df) 

# Check the resulting episode structure 
check_eps(episode_df)
```

# Step 3: Describe treatment episodes

Once episodes have been identified, each unique episode can be summarized using `describe_episodes()` function shown below

```{r}
describe_episodes(episode_df)
```

These summaries provide an initial description of treatment utilization and the organization of repeated episodes within clients.

# Step 4: Visualize episodes

Before fitting statistical models, it is often useful to visually inspect patterns of change within and across treatment episodes. The `plot_` family of functions provides several ways to explore these patterns. For example, `plot_episode_curves()` displays individual client trajectories alongside the average linear pattern within each episode.

```{r, episode-curves, fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
plot_episode_curves(episode_df)
```

Visualizing treatment trajectories can reveal differences in starting levels, treatment duration, rates of change, and variability across successive episodes. Because the clients contributing to each episode number may differ, these plots are primarily descriptive and should not be interpreted as adjusted within-client effects. For additional plotting options see [Visualizing Treatment Episodes](articles/viz-episodes.html)

# Step 5: Model treatment outcomes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care. These methods differ in whether change is modeled directly from session-level observations or summarized at the episode level.

For example, a longitudinal mixed-effects model can be fit directly to session-level data:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

This approach estimates change within treatment episodes while accounting for the hierarchical structure of sessions nested within episodes and clients.

Alternative functions support slopes-as-outcomes and Bayesian multilevel approaches for examining change within and across episodes. For a detailed discussion of the available models and their interpretation, see [Modeling Treatment Episodes](articles/modeling-episodes.html)
