---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Basic workflow

A typical `leap` workflow consists of identifying unique treatment episodes in longitudinal health records data, describing the resulting episode structure, and then visualizing or modeling change across episodes.

```{r data, include = FALSE, eval = FALSE}

# package 
library(tidyverse)

# create a small example data set.
load("data/simulation_data.RData")

# subset simulated data 
data <- data %>%
  pluck("unbalanced", "clinical", "diminishing_resp") %>%
  nest_by(client_id) %>%
  ungroup() %>%
  slice_head(n = 10) %>%
  unnest(cols = data)

```

### Episode identification

```{r identify, eval = FALSE}
# Data check 
check_raw(data)

# Derive episodes from raw 
episode_data <- add_episode_vars(data)

# Check the episode structure 
check_eps(episode_data)
```

### Describe episodes

```{r describe, eval = FALSE}
describe_episodes(episode_data)
```

### Model change within and between episodes

```{r model, eval = FALSE}
# mixed-effects growth model 
fit_lme(episode_data, cohort = "all", model = "episode")
```

### Visualize treatment trajectories

```{r visualize, eval = FALSE}
# plot growth rates by episode 
plot_episode_curves(episode_data)
```

## Episode identification

Treatment episodes are commonly defined using a predetermined period of inactivity, such as 90 days without treatment. `leap` supports this approach while also providing tools for examining the empirical distribution of time between sessions.

Session lags can be calculated and inspected directly:

```{r lags, eval = FALSE}
episode_data <- data |>
  add_session_lag() |>
  add_episode_id(delimiter = 90)
```

Candidate episode delimiters can also be estimated from the observed distribution of session lags:

```{r delimiter, eval = FALSE}
lag_delimiter(data, method = "iqr")
lag_delimiter(data, method = "quantile")
lag_delimiter(data, method = "mixture")
```

The mixture-modeling approach, implemented using the `mclust` package [@scrucca2016], treats episode identification as an empirical classification problem by identifying relatively short- and long-gap components in the observed distribution of session lags. Estimated delimiters should be interpreted as data-informed candidate thresholds rather than definitive clinical boundaries between treatment episodes.

## Modeling treatment episodes

`leap` currently supports several complementary approaches to analyzing change across repeated treatment episodes:

- `fit_lme()` fits longitudinal mixed-effects models directly to session-level observations.
- `fit_sao()` uses episode-specific rates of change as outcomes in a slopes-as-outcomes framework.
- `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using `lme4` [@bates2015] whereas Bayesian multilevel models are estimated using `brms` [@burkner2017]. These approaches address related but distinct questions about therapeutic change and the structure of repeated episodes of care.

`leap` includes simulated psychotherapy data for demonstrating episode identification, visualization, and statistical modeling. The simulated data illustrate common features of longitudinal treatment records, including unequal numbers of sessions, repeated episodes of care, and variation in within- and between-episode change.
