---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical `leap` workflow begins with longitudinal treatment data in which each row represents a treatment session. From these data, `leap` provides functions for identifying distinct episodes of care, describing the resulting episode structure, visualizing treatment trajectories.

This guide provides a brief introduction to that workflow. More detailed discussions of each step are available in the package articles.

# Data structure

To utilize all that is needed is a person-period longitudinal data frame. Below is an example from `leaps` built-in data.

```{r data, include = FALSE}

# package 
# library(leap)
library(tidyverse)

# load example data
load("~/Library/CloudStorage/Dropbox/01_desk/04_personal/02_data_science/02_projects/01_pkgs/leap/data/simulation_data.RData")

# select subset for vignette 
raw_df <- data %>%
  pluck("unbalanced", "clinical", "diminishing_resp") %>%
  nest_by(client_id) %>%
  ungroup() %>%
  slice_head(n = 10) %>%
  unnest(cols = data)

```

```{r}
head(raw_df)
```

# Identify Treatment Episodes

From these raw data, data can be checked using the `check_` functions. Subsequently, the required variables to perform episode analyses can be calculated using the `add_` functions. These variables can derived one-by-one, or all at once as shown below:

```{r identify, eval = FALSE}
# Data check 
check_raw(data)

# Derive episodes from raw 
episode_data <- add_episode_vars(data)

# Check the episode structure 
check_eps(episode_data)
```

# Describe episodes

Once episodes have been identified, each unique episode can be summarized using `describe_episodes()` function shown below

```{r describe, eval = FALSE}
describe_episodes(episode_data)
```

These summaries provide an initial description of treatment utilization and the organization of repeated episodes within clients.

# Visualize treatment trajectories

Episode-specific trajectories can be inspected before fitting statistical models:

```{r visualize, eval = FALSE}
# plot growth rates by episode 
plot_episode_curves(episode_data)
```

Visualizing the data can help reveal differences in treatment duration, starting levels, rates of change, and patterns of return across episodes.

# Model change within and between episodes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care.

A longitudinal mixed-effects model can be fit directly to session-level observations:

```{r model, eval = FALSE}
# mixed-effects growth model 
fit_lme(episode_data, cohort = "all", model = "episode")
```

Additional modeling functions support alternative representations of episode-level change:

- `fit_lme()` fits longitudinal mixed-effects models to session-level observations.
- `fit_sao()` models episode-specific rates of change using a slopes-as-outcomes framework.
- `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using lme4 [@bates2015], whereas Bayesian multilevel models are estimated using brms [@burkner2017]. These approaches address related but distinct questions about change within and across repeated episodes of care

# Modeling treatment episodes

`leap` currently supports several complementary approaches to analyzing change across repeated treatment episodes:

- `fit_lme()` fits longitudinal mixed-effects models directly to session-level observations.
- `fit_sao()` uses episode-specific rates of change as outcomes in a slopes-as-outcomes framework.
- `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using `lme4` [@bates2015] whereas Bayesian multilevel models are estimated using `brms` [@burkner2017]. These approaches address related but distinct questions about therapeutic change and the structure of repeated episodes of care.

`leap` includes simulated psychotherapy data for demonstrating episode identification, visualization, and statistical modeling. The simulated data illustrate common features of longitudinal treatment records, including unequal numbers of sessions, repeated episodes of care, and variation in within- and between-episode change.
