---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical leap workflow begins with longitudinal treatment data in which each row represents a treatment session. From these data, leap provides tools for identifying distinct episodes of care, describing the resulting episode structure, visualizing treatment trajectories, and modeling change within and across episodes.

This guide provides a brief introduction to that workflow. More detailed discussions of each step are available in the package articles.

# Example data

leap includes simulated longitudinal psychotherapy data that illustrate common features of real-world treatment records, including unequal numbers of sessions, repeated episodes of care, and variation in treatment trajectories.

```{r data, include = FALSE, eval = FALSE}

# package 
library(leap)
library(tidyverse)

# load example data
load("data/simulation_data.RData")

# select subset for vignette 
data <- data %>%
  pluck("unbalanced", "clinical", "diminishing_resp") %>%
  nest_by(client_id) %>%
  ungroup() %>%
  slice_head(n = 10) %>%
  unnest(cols = data)

```

# Identify Treatment Episodes

```{r identify, eval = FALSE}
# Data check 
check_raw(data)

# Derive episodes from raw 
episode_data <- add_episode_vars(data)

# Check the episode structure 
check_eps(episode_data)
```

# Describe episodes

Once episodes have been identified, each unique episode can be summarized using `describe_episodes()`

```{r describe, eval = FALSE}
describe_episodes(episode_data)
```

These summaries provide an initial description of treatment utilization and the organization of repeated episodes within clients.

# Visualize treatment trajectories

Episode-specific trajectories can be inspected before fitting statistical models:

```{r visualize, eval = FALSE}
# plot growth rates by episode 
plot_episode_curves(episode_data)
```

Visualizing the data can help reveal differences in treatment duration, starting levels, rates of change, and patterns of return across episodes.

# Model change within and between episodes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care.

A longitudinal mixed-effects model can be fit directly to session-level observations:

```{r model, eval = FALSE}
# mixed-effects growth model 
fit_lme(episode_data, cohort = "all", model = "episode")
```

Additional modeling functions support alternative representations of episode-level change: - `fit_lme()` fits longitudinal mixed-effects models to session-level observations. - `fit_sao()` models episode-specific rates of change using a slopes-as-outcomes framework. - `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using lme4 [@bates2015], whereas Bayesian multilevel models are estimated using brms [@burkner2017]. These approaches address related but distinct questions about change within and across repeated episodes of care

# Episode identification

Treatment episodes are commonly defined using a predetermined period of inactivity, such as 90 days without treatment. `leap` supports this approach while also providing tools for examining the empirical distribution of time between sessions.

Session lags can be calculated and inspected directly:

```{r lags, eval = FALSE}
episode_data <- data |>
  add_session_lag() |>
  add_episode_id(delimiter = 90)
```

Candidate episode delimiters can also be estimated from the observed distribution of session lags:

```{r delimiter, eval = FALSE}
lag_delimiter(data, method = "iqr")
lag_delimiter(data, method = "quantile")
lag_delimiter(data, method = "mixture")
```

The mixture-modeling approach, implemented using the `mclust` package [@scrucca2016], treats episode identification as an empirical classification problem by identifying relatively short- and long-gap components in the observed distribution of session lags. Estimated delimiters should be interpreted as data-informed candidate thresholds rather than definitive clinical boundaries between treatment episodes.

# Modeling treatment episodes

`leap` currently supports several complementary approaches to analyzing change across repeated treatment episodes:

- `fit_lme()` fits longitudinal mixed-effects models directly to session-level observations.
- `fit_sao()` uses episode-specific rates of change as outcomes in a slopes-as-outcomes framework.
- `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using `lme4` [@bates2015] whereas Bayesian multilevel models are estimated using `brms` [@burkner2017]. These approaches address related but distinct questions about therapeutic change and the structure of repeated episodes of care.

`leap` includes simulated psychotherapy data for demonstrating episode identification, visualization, and statistical modeling. The simulated data illustrate common features of longitudinal treatment records, including unequal numbers of sessions, repeated episodes of care, and variation in within- and between-episode change.
