---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical `leap` workflow begins with longitudinal treatment data in which each row represents a treatment session for a client. From these records, `leap` provides tools for identifying distinct episodes of care, describing and visualizing the resulting episode structure, and analyzing change within and across treatment episodes.

This guide provides a brief introduction to the core `leap` workflow. More detailed discussions of episode identification, visualization, and statistical modeling are available in the package articles.

# Step 1: Check data

To get started, `leap` requires person-period longitudinal data, with one row representing each treatment session. At minimum, the data should contain variables identifying the client, session, session date, and outcome of interest.

The example below uses a small data set included with leap:

```{r data, include = FALSE}

# package 
# library(leap)
library(tidyverse)

# load example data
load("~/Library/CloudStorage/Dropbox/01_desk/04_personal/02_data_science/02_projects/01_pkgs/leap/data/simulation_data.RData")

# select subset for vignette 
raw_df <- data %>%
  pluck("unbalanced", "clinical", "diminishing_resp") %>%
  nest_by(client_id) %>%
  ungroup() %>%
  slice_head(n = 10) %>%
  unnest(cols = data)

```

```{r}
head(raw_df)
```

`leap` uses a standardized set of variable names throughout its functions. If the original data use different names, `cols_standard()` can be used to rename the core variables before proceeding:

```{r, eval = FALSE}
data <- cols_standard(
  data = raw_df, 
  client = "case_id", 
  session = "session_num", 
  date = "session_date",
  outcome = "depression_score"
  )
```

Once the variables have been standardized, `check_raw()` can be used to verify that the data contain the information required to identify treatment episodes:

```{r, eval = FALSE}
check_raw(data)
```

This initial check helps identify missing or improperly structured variables before episode-level variables are derived.

# Step 2: Identify Treatment Episodes

Once the session-level data have been prepared, leap can be used to identify distinct treatment episodes and derive the variables needed for subsequent episode analyses. The check\_ functions evaluate the structure of the data before and after episode identification, while the add\_ functions derive treatment episode variables.

Episode variables can be added individually or created together using add_episode_vars(). The complete workflow is shown below::

```{r identify, eval = FALSE}
# Check data prior to episode identification 
check_raw(data) 

# Identify treatment episodes and derive episode variables
episode_data <- add_episode_vars(data) 

# Check the resulting episode structure 
check_eps(episode_data)
```

# Step 3: Describe episodes

Once episodes have been identified, each unique episode can be summarized using `describe_episodes()` function shown below

```{r describe, eval = FALSE}
describe_episodes(episode_data)
```

These summaries provide an initial description of treatment utilization and the organization of repeated episodes within clients.

# Step 4: Visualize treatment trajectories

Episode-specific trajectories can be inspected before fitting statistical models:

```{r visualize, eval = FALSE}
plot_episode_curves(episode_data)
```

Visualizing the data can help reveal differences in treatment duration, starting levels, rates of change, and patterns of return across episodes. For more see [Visualizing Treatment Episodes](articles/viz-episodes.html)

# Step 5: Model outcomes within and between episodes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care.

A longitudinal mixed-effects model can be fit directly to session-level observations:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

Additional modeling functions support alternative representations of episode-level change, see [Modeling Treatment Episodes](articles/modeling-episodes.html)
