---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical `leap` workflow progresses from preparing the raw data to identifying, describing, visualizing, and modeling repeated episodes of treatment. The package is designed to accommodate real-world treatment records in which clients may attend different numbers of sessions and episodes of care, with some completing a single episode and others returning for treatment multiple times over time.

# Step 1: Prepare and check

`leap` works with person-period longitudinal data, where each row represents a treatment session for a client. At minimum, the data should include variables identifying the `client`, `session`, `session date`, and the `outcome` of interest measured at each session.

The example below uses a small data set included with the `leap` package:

```{r setup, include = FALSE}

# packages
library(tidyverse)
devtools::load_all()

# load example data
load("~/Library/CloudStorage/Dropbox/01_desk/04_personal/02_data_science/02_projects/01_pkgs/leap/data/simulation_data.RData")

# select subset for vignette 
raw_df <- data %>%
  purrr::pluck("unbalanced", "clinical", "diminishing_resp") %>%
  dplyr::nest_by(client_id) %>%
  dplyr::ungroup() %>%
  dplyr::slice_head(n = 10) %>%
  tidyr::unnest(cols = data)

```

```{r}
head(raw_df)
```

It is important to be aware that `leap` uses a standardized set of variable names throughout its functions. If the original data use different names, `cols_standard()` can be used to rename the core variables before proceeding.

Once the variables have been standardized, `check_raw()` can be used to verify that the data contain the information required to identify treatment episodes:

```{r, check raw data}
check_raw(raw_df)
```

The output from this initial check identifies the variables already present in the data, the episode-related variables that still need to be created, and the corresponding `add_*()` function for creating each of the missing variables.

# Step 2: Identify episodes

Once the session-level data have been prepared, `leap` can be used to identify distinct treatment episodes and derive the variables needed for subsequent episode analyses. The `check_` functions evaluate the structure of the data before and after episode identification, while the `add_` functions create episode indicator variables.

The necessary episode variables can be added individually or created together using the wrapper `add_episode_vars()`. The complete workflow is shown below:

```{r identify}
# create all episode variables
episode_df <- add_episode_vars(raw_df) 

# Check episode structure 
check_eps(episode_df)
```

# Step 3: Describe episodes

Once episode indicators have been added, each unique episode can be summarized using `describe_episodes()` function shown below

```{r}
describe_episodes(episode_df)
```

This summary provides an initial description of service utilization and the organization of repeated episodes within clients.

# Step 4: Visualize episodes

Before fitting statistical models, it is often useful to visually inspect patterns of change within and across treatment episodes. The `plot_` family of functions provides several ways to explore these patterns. For example, `plot_episode_curves()` displays individual client trajectories alongside the average linear pattern within each episode.

```{r, episode-curves, fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
plot_episode_curves(episode_df)
```

Visualizing treatment trajectories can reveal differences in starting levels, treatment duration, rates of change, and variability across successive episodes. Because the clients contributing to each episode number may differ, these plots are primarily descriptive and should not be interpreted as adjusted within-client effects. For additional plotting options see [Visualizing Treatment Episodes](articles/viz-episodes.html)

# Step 5: Model outcomes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care. These methods differ in whether change is modeled directly from session-level observations or summarized at the episode level.

For example, a longitudinal mixed-effects model can be fit directly to session-level data:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

This approach estimates change within treatment episodes while accounting for the hierarchical structure of sessions nested within episodes and clients.

Alternative functions support slopes-as-outcomes and Bayesian multilevel approaches for examining change within and across episodes. For a detailed discussion of the available models and their interpretation, see [Modeling Treatment Episodes](articles/modeling-episodes.html)
