---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A typical `leap` workflow begins with longitudinal treatment data in which each row represents a treatment session. From these data, `leap` provides functions for identifying distinct episodes of care, describing the resulting episode structure, visualizing treatment trajectories.

This guide provides a brief introduction to that workflow. More detailed discussions of each step are available in the package articles.

# Data Structure

To utilize all that is needed is a person-period longitudinal data frame. Below is an example from `leaps` built-in data.

```{r data, include = FALSE}

# package 
# library(leap)
library(tidyverse)

# load example data
load("~/Library/CloudStorage/Dropbox/01_desk/04_personal/02_data_science/02_projects/01_pkgs/leap/data/simulation_data.RData")

# select subset for vignette 
raw_df <- data %>%
  pluck("unbalanced", "clinical", "diminishing_resp") %>%
  nest_by(client_id) %>%
  ungroup() %>%
  slice_head(n = 10) %>%
  unnest(cols = data)

```

```{r}
head(raw_df)
```

# Step 1: Identify Treatment Episodes

From these raw data, data can be checked using the `check_` functions. Subsequently, the required variables to perform episode analyses can be calculated using the `add_` functions. These variables can derived one-by-one, or all at once as shown below:

```{r identify, eval = FALSE}
check_raw(data)

episode_data <- add_episode_vars(data)

check_eps(episode_data)
```

## Step 2: Describe episodes

Once episodes have been identified, each unique episode can be summarized using `describe_episodes()` function shown below

```{r describe, eval = FALSE}
describe_episodes(episode_data)
```

These summaries provide an initial description of treatment utilization and the organization of repeated episodes within clients.

## Step 3: Visualize treatment trajectories

Episode-specific trajectories can be inspected before fitting statistical models:

```{r visualize, eval = FALSE}
plot_episode_curves(episode_data)
```

Visualizing the data can help reveal differences in treatment duration, starting levels, rates of change, and patterns of return across episodes. For more see [Visualizing Treatment Episodes](articles/viz-episodes.html)

## Step 4: Model outcomes within and between episodes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care.

A longitudinal mixed-effects model can be fit directly to session-level observations:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

Additional modeling functions support alternative representations of episode-level change, see [Modeling Treatment Episodes](articles/modeling-episodes.html)
