---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

This article introduces a typical `leap` workflow, progressing from preparing raw data to identifying, describing, visualizing, and modeling repeated episodes of treatment.

# Prepare data

`leap` works with longitudinal data frames, in which each row represents a treatment session for one client. At minimum, the raw data should include a core set of variables identifying the client, session, session date, along with one or more outcomes measured during treatment.

The example below uses simulated data included with `leap` to illustrate the expected data structure:

```{r setup, include = FALSE}

# use leap package 
devtools::load_all()

# Load simulated data from leap 
data("simulation_data", package = "leap") 

# Select a small subset for the vignette 
raw_df <- simulation_data$unbalanced$clinical$diminishing_resp

```

```{r}
head(raw_df)
```

It is important to be aware that `leap` uses a standardized set of variable names throughout its functions. If the original data use different names, `cols_standard()` can be used to rename the core variables before proceeding.

Once the variables have been standardized, `check_raw()` can be used to verify that the data contain the information required to identify treatment episodes:

```{r, check raw data}
check_raw(raw_df)
```

The output from this initial check identifies the variables already present in the data, the episode-related variables that still need to be created, and the corresponding `add_*()` function to create the new variable.

# Add variables

Once the session-level data have been prepared, the `add_*()` family of functions can be used to generate the variables needed to represent treatment episodes in the data. These additional variables are:

- session_lag: number of days elapsed since the previous session
- episode_id: identifies each distinct treatment episode within a client
- episode_session: identifies the consecutive session number within each treatment episode
- n_episodes: indicates the total number of treatment episodes for each client
- client_episode_id: provides a unique identifier for each client-episode combination

Each variable can be added individually using their respective `add_*()` functions or created simultaneously using the `add_episode_vars()` wrapper function:

```{r, adding episode variables}
# add variables simultaneously 
episode_df <- add_episode_vars(raw_df) 
```

# Summarize episodes

After all episode variables have been added, it is useful to examine the resulting data before proceeding. The `check_episodes()` function provides an overview of the episode structure and evaluates several potential data issues, including total number of observations, unique clients, unique treatment episodes, the number of treatment episodes, session and chronological ordering, and missing values.

```{r, check episodes, warning = FALSE}
check_episodes(episode_df)
```

Importantly, `check_episodes()` also warns users when treatment episodes contain too few observations to estimate change or when the number of observations may result in unreliable estimates.

Treatment episodes can be further summarized using the `describe_episodes()` function. Whereas `check_episodes()` focuses primarily on data structure and potential issues, `describe_episodes()` provides descriptive summaries of service utilization and the characteristics of each treatment episode represented in the data.

```{r}
describe_episodes(episode_df)
```

# Visualize episodes

Before fitting statistical models, it is often useful to visually inspect patterns of change within and across treatment episodes. The `plot_*()` family of functions provides several ways to explore these patterns. For example, `plot_episode_curves()` displays individual client growth curves alongside the average linear growth curve within each episode.

```{r, episode-curves, fig.width=7, fig.height=5, fig.align="center"}
plot_episode_curves(episode_df)
```

Visualizing each episode growth curve can reveal differences in starting levels, episode duration, rates of change, and variability between different episodes. Because the clients contributing to each episode number may differ, these plots are primarily descriptive and should not be interpreted as adjusted within-client effects.

# Model outcomes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care. These methods differ in whether change is modeled directly from session-level observations or summarized at the episode level.

For example, a longitudinal mixed-effects model can be fit directly to session-level data:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

This approach estimates change within treatment episodes while accounting for the hierarchical structure of sessions nested within episodes and clients.

Alternative functions support slopes-as-outcomes and Bayesian multilevel approaches for examining change within and across episodes. For a detailed discussion of the available models and their interpretation see [Model Episodes](articles/modeling-episodes.html) article.

# Summary

This article introduced a typical `leap` workflow. The package functions offer considerably more flexibility than can be demonstrated here. For additional options and guidance, consult the function documentation and the subsequent articles.

- [Define Episodes](articles/define-episodes.html)
- [Visualize Episodes](articles/viz-episodes.html)
- [Model Episodes](articles/modeling-episodes.html)
- [Research Foundations](articles/rx-foundations.html)
