---
title: "Get Started with leap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with leap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

A typical `leap` workflow progresses from preparing the raw data to identifying, describing, visualizing, and modeling repeated episodes of treatment. The package is designed to accommodate real-world treatment records in which clients may attend different numbers of sessions and episodes of care, with some completing a single episode and others returning for treatment multiple times over time.

# Step 1: Prepare and check

`leap` works with person-period longitudinal data, where each row represents a treatment session for a client. At minimum, the raw data should include variables identifying the `client`, `session`, `session date`, and the `outcome` of interest measured at each session.

The example below illustrates how the data can look:

```{r setup, include = FALSE}

# use leap package 
devtools::load_all()

# Load simulated data from leap 
data("simulation_data", package = "leap") 

# Select a small subset for the vignette 
raw_df <- simulation_data$unbalanced$clinical$diminishing_resp

```

```{r}
head(raw_df)
```

It is important to be aware that `leap` uses a standardized set of variable names throughout its functions. If the original data use different names, `cols_standard()` can be used to rename the core variables before proceeding.

Once the variables have been standardized, `check_raw()` can be used to verify that the data contain the information required to identify treatment episodes:

```{r, check raw data}
check_raw(raw_df)
```

The output from this initial check identifies the variables already present in the data, the episode-related variables that still need to be created, and the corresponding `add_*()` function to create each of the missing variables.

# Step 2: Identify episodes

Once the session-level data have been prepared, leap can identify distinct treatment episodes and generate the episode-level variables required for subsequent analyses. By default, `leap` demarcates episodes based on periods of treatment inactivity, specifically the number of days between consecutive sessions. See [Define Episodes](articles/define-episodes.html) for additional details.

Episode indicators can be added individually or simultaneously using the wrapper function `add_episode_vars()`. The new variables this function creates are

- `session_lag`: identifies the number of days elapsed since last session
- `episode_id`: identifies each distinct treatment episode for a client
- `episode_session`: identifies the consecutive session number in each episode
- `n_episodes`: indicates the total number of treatment episodes a client attended

```{r, adding episode variables}
# create all episode variables
episode_df <- add_episode_vars(raw_df) 
```

After all episode variables are added, the `check_eps()` function can be used to check the resulting data structure.

```{r, check episode variables}
# check episode structure 
check_eps(episode_df)
```

Importantly, `check_eps()` will also warn the user when episodes contain too few observations to reliably estimate change.

# Step 3: Describe episodes

Once episode indicators have been added, each unique episode can be summarized using `describe_episodes()` function shown below

```{r}
describe_episodes(episode_df)
```

This summary provides an initial description of service utilization and the characteristics of each treatment episode contained in the data.

# Step 4: Visualize episodes

Before fitting statistical models, it is often useful to visually inspect patterns of change within and across treatment episodes. The `plot_*()` family of functions provides several ways to explore these patterns. For example, `plot_episode_curves()` displays individual client trajectories alongside the average linear pattern within each episode.

```{r, episode-curves, fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
plot_episode_curves(episode_df)
```

Visualizing treatment trajectories can reveal differences in starting levels, treatment duration, rates of change, and variability across successive episodes. Because the clients contributing to each episode number may differ, these plots are primarily descriptive and should not be interpreted as adjusted within-client effects. For additional plotting options see [Visualize Episodes](articles/viz-episodes.html)

# Step 5: Model outcomes

`leap` provides several complementary approaches for modeling therapeutic change across repeated episodes of care. These methods differ in whether change is modeled directly from session-level observations or summarized at the episode level.

For example, a longitudinal mixed-effects model can be fit directly to session-level data:

```{r model, eval = FALSE}
mod <- fit_lme(episode_data, cohort = "all", model = "episode")
```

This approach estimates change within treatment episodes while accounting for the hierarchical structure of sessions nested within episodes and clients.

Alternative functions support slopes-as-outcomes and Bayesian multilevel approaches for examining change within and across episodes. For a detailed discussion of the available models and their interpretation, see [Model Episodes](articles/modeling-episodes.html)

# Summary

This article introduced a typical leap workflow. The package functions offer considerably more flexibility than can be demonstrated here. For additional options and guidance, consult the function documentation and the subsequent articles.
