---
title: "Model Episodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

`leap` currently supports several complementary approaches:

- `fit_lme()` fits longitudinal mixed-effects models directly to session-level observations.
- `fit_sao()` uses episode-specific rates of change as outcomes in a slopes-as-outcomes framework.
- `fit_brms()` fits Bayesian multilevel models of session-level treatment trajectories.

Frequentist mixed-effects models are estimated using `lme4` whereas Bayesian multilevel models are estimated using `brms`. These approaches address related but distinct questions about therapeutic change and the structure of repeated episodes of care.

# Levels of analysis

An important distinction among the modeling approaches in `leap` is the level at which change is analyzed. The **session-level models** retain each treatment session as the level-1 unit. This allows change over the course of treatment to be modeled directly while accounting for sessions belonging to the same episode and multiple episodes belonging to the same client. Both `fit_lme()` and `fit_brms()` utilize this approach.

By contrast, **episode-level models** first summarize change within each treatment episode using a measure such as a pre-post change score or estimated rate of change. For example, a linear regression slope can be estimated for each client-episode to represent the rate of change across sessions. These episode-specific slopes can then serve as the outcome in a multilevel model where the level-1 unit is the episode of treatment. This slopes-as-outcomes approach is implemented in `fit_sao()`.

The most appropriate model depends on the research question and aims of the analysis.

# Longitudinal mixed-effects

`fit_lme()` fits a longitudinal mixed-effects model to the data with `lme4`. For example, a basic growth model can be fit using:

```{r, eval = FALSE}
fit_lme(episode_data, model = "session")
```

In this particular model specification, the corresponding three-level equation is:

$$Y_{ijk} = \beta_0 + \beta_1(\text{Session}{ijk}) + \left[u{0k} + u_{1k}(\text{Session}{ijk}) + v{0jk} + v_{1jk}(\text{Session}{ijk}) \right] + \varepsilon{ijk}$$

where $Y_{ijk}$ is the outcome at session $i$ in episode $j$ for client $k$. The fixed effects $\beta_0$ and $\beta_1$ represent the average starting level and rate of change across sessions, respectively. The terms in brackets represent random effects, allowing starting levels and rates of change to vary across clients ($u_{0k}$ and $u_{1k}$) and across episodes within clients ($v_{0jk}$ and $v_{1jk}$). The residual $\varepsilon_{ijk}$ represents remaining session-level variation.

Episode number can be added as a fixed predictor to examine whether outcomes systematically differ across successive episodes:

```{r, eval = FALSE}
fit_lme(episode_data, cohort = "all", model = "episode")
```

The available model specifications provide a progression from describing variation in outcomes to examining change within and across treatment episodes:

| Model | Purpose |
|:--:|:---|
| "null" | Describe variation across clients and episodes without modeling change over time |
| "session" | Estimate change across sessions within treatment episodes |
| "episode" | Estimate within-episode change while accounting for episode number |
| "full" | Examine whether within-episode change differs across successive episodes |

The "full" model includes the interaction between session and episode number. This interaction addresses a particularly important multi-episode question: does the rate of change during treatment differ depending on which episode of care a client is attending?

For example, a positive or negative interaction may indicate that treatment trajectories become steeper or flatter across successive episodes, depending on the direction and scaling of the outcome measure.

# Slopes-as-outcomes

An alternative approach is to summarize change separately within each treatment episode and then analyze those estimates at the episode level. `fit_sao()` implements this slopes-as-outcomes approach. Conceptually, the analysis occurs in two stages. First, a OLS regression slope is estimated for each client in each episode using the `episode_slopes()` function. Second, those episode-specific slopes are modeled as outcomes using `fit_sao()` function.

This two step process can be implemented as follows:

```{r, eval = FALSE}
# get slopes 
slopes_df <- episode_slopes(raw_data)

# fit sao model 
fit_sao(slopes_df, model = "session")
```

For the "session" specification, the second-stage model can be written as:

$$Y_{ij} = \beta_{0} + \beta_{10}(Sessions_{ij}) + [U_{0j} + e_{ij}]$$

where $Y_{ij}$ is the estimated treatment slope in episode *i* for client *j*. The fixed effect $\beta_{0}$ represents the expected episode-specific rate of change, while $\beta_{10}$ represents the association between the total number of sessions in an episode and the estimated rate of change. The random intercept $U_{0j}$ accounts for differences in average rates of change across clients, and $e_{ij}$ represents remaining variation among episode-specific slopes within clients.

Including the number of sessions is particularly relevant in a slopes-as-outcomes analysis because episodes may differ considerably in length. The `"session"` model therefore examines whether estimated rates of change systematically vary with the number of sessions contributing to each episode.

Other model specifications extend this framework to examine whether rates of change differ across successive treatment episodes. For example:

```{r, eval = FALSE}
fit_sao( slopes_df, model = "episode" )
```

adds episode number as a predictor, allowing the model to examine whether episode-specific rates of change systematically differ across repeated episodes of care.

# Bayesian multilevel models

`fit_brms()` provides a Bayesian approach to modeling session-level treatment trajectories using `brms`. The general model structure parallels the longitudinal mixed-effects models available through `fit_lme()`:

```{r, eval = FALSE}
fit_brms(episode_data, model = "episode")
```

For the `"episode"` specification, the outcome model can be written as:

$$\beta_0 + \beta_1(\text{Session}{ijk}) + \beta_2(\text{Episode}{jk}) + \left[u_{0k} + u_{1k}(\text{Session}{ijk}) + v{0jk} + v_{1jk}(\text{Session}{ijk}) \right] + \varepsilon{ijk}$$

where $Y_{ijk}$ is the outcome at session $i$ in episode $j$ for client $k$. The fixed effects $\beta_0$, $\beta_1$, and $\beta_2$ represent the average starting level, rate of change across sessions, and association between episode number and the outcome, respectively. The terms in brackets represent random effects that allow starting levels and rates of change to vary across clients and across episodes within clients.

In a Bayesian model, these parameters are treated as unknown quantities described by probability distributions. For example, the model can be completed by specifying: \$\$

$$\varepsilon_{ijk} \sim \mathrm{N}(0,\sigma^2),$$

together with prior distributions such as

$$\beta_p \sim \mathrm{N} (0,\tau_\beta^2),$$

and multivariate normal distributions for the client- and episode-level random effects:

$$\begin{pmatrix} u_{0k} \ u_{1k} \end{pmatrix} \sim \mathrm{N} \left( \begin{pmatrix} 0 \ 0 \end{pmatrix}, \Sigma_{\text{client}} \right),$$

$$\begin{pmatrix} v_{0jk} \ v_{1jk} \end{pmatrix} \sim \mathrm{N} \left(\begin{pmatrix} 0 \ 0 \end{pmatrix}, \Sigma_{\text{episode}} \right).$$

The primary distinction from `fit_lme()` is therefore not the structure of the treatment trajectory itself, but how uncertainty about the model parameters is represented and estimated. Rather than producing point estimates and sampling-based confidence intervals, Bayesian estimation combines the likelihood with prior distributions to obtain posterior distributions for the model parameters.

These posterior distributions can be used to summarize the most plausible values of treatment effects, quantify uncertainty using credible intervals, and calculate probabilities for quantities of substantive interest.

Because Bayesian multilevel models can require substantially more computation than frequentist mixed-effects models, `fit_lme()` can provide a useful starting point for evaluating model specification before fitting an analogous model with `fit_brms()`.

# Cohort selection

Not every research question requires the same set of clients. In particular, analyses of repeated treatment episodes may distinguish between the full sample versus clients who only contributed multiple episodes of care.

All of the `fit_*()` include a cohort argument which provides a consistent way to specify the analytic sample desired for analysis:

```{r, eval = FALSE}
fit_lme(episode_data, cohort = "multiple", model = "episode")
```

Using cohort = "multiple" restricts the analysis to clients who contributed more than one treatment episode. This can be useful when the research question specifically concerns patterns of change across repeated episodes.

In contrast, cohort = "all" retains all eligible clients, including those observed for only one episode. The choice should be guided by the research question. Restricting analyses to clients who return for treatment changes the population being analyzed and should therefore be considered when interpreting results.

# Centering predictors

The interpretation of model coefficients depends partly on how session and episode number are coded.

By default, leap can center these variables so that the first treatment session and first treatment episode provide meaningful reference points. Under this parameterization, the model intercept corresponds more closely to expected outcome levels at the beginning of the initial episode of care.

Centering can be controlled using the center argument:

```{r, eval = FALSE}
fit_lme(episode_data, center = TRUE, cohort = "all", model = "episode")
```

Centering does not change the underlying observations. Rather, it changes the reference point used to interpret model coefficients and can also improve the numerical behavior of models containing random slopes.
