---
title: "Visualize Episodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualize Episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

Visualizing episodes can help researchers understand when episodes occur, how outcomes change within episodes, and what happens during periods between episodes.

`leap` provides plotting functions for several complementary features of multi-episode data:

- `plot_lag_density()` visualizes the distribution of time between consecutive sessions.
- `plot_episode_curves()` displays session-level outcome trajectories within successive treatment episodes.
- `plot_episode_change()` summarizes treatment change across episode number.
- `plot_breaks_loss()` visualizes outcome loss between consecutive treatment episodes.

These functions operate at different levels of the treatment record. Some use session-level observations directly, whereas others use episode- or transition-level summaries produced by other `leap` functions.

```{r, load data, include = FALSE}

library(ggplot2)
library(scales)

# leap functions
devtools::load_all()

# simulated data 
data("simulation_data", package = "leap") 

# select one dataframe
episode_data <- simulation_data$unbalanced$clinical$diminishing_resp
episode_data <- add_episode_vars(data = episode_data)


```

# Plot session lags

Treatment episodes are commonly distinguished by the amount of elapsed time between consecutive sessions. Before selecting an episode delimiter, it can be useful to inspect the empirical distribution of session lags.

```{r, lag-density, warning=FALSE, fig.width=6, fig.height=4, fig.align="center"}
plot_lag_density(episode_data)
```

`plot_lag_density()` displays the density of positive `session_lag` values. The first session observed for each client has an undefined lag and is excluded from the distribution.

A candidate delimiter can also be displayed directly:

```{r, lag-density-delimiter, warning = FALSE, fig.width=6, fig.height=4, fig.align="center"}
plot_lag_density(episode_data, delimiter = 90)
```

# Plot episode growth curves

Once episodes have been identified, `plot_episode_curves()` provides a direct view of session-level change within successive episodes.

```{r episode-curves, warning = FALSE, fig.width=7, fig.height=4, fig.align="center"}
plot_episode_curves(episode_data)
```

Each panel represents a treatment episode number. Individual client-episode trajectories are displayed in the background, while a pooled linear trend is superimposed within each episode.

For example, the Episode 2 panel contains observations from clients who contributed a second treatment episode. Because fewer clients typically contribute observations to later episodes, the composition of the sample may differ across panels.

The plot is therefore primarily descriptive. Differences between episode-specific trends should not automatically be interpreted as adjusted within-client effects.

```{r episode-curves-no-cutoff, warning = FALSE, fig.width=7, fig.height=4, fig.align="center"}
plot_episode_curves(episode_data, clinical_cutoff = NULL)
```

`plot_episode_curves()` is particularly useful during exploratory analysis because it makes several features of the data visible simultaneously, including variability in baseline scores, rates of change, episode length, and the number of clients contributing to successive episodes.

# Plot episode change scores

Session-level trajectories can also be reduced to episode-level summaries using `episode_slopes()`.

```{r episode-slopes, warning = FALSE, fig.width=5, fig.height=4, fig.align="center"}
episode_summary <- episode_slopes(episode_data, higher_is_better = TRUE)
```

The returned data contain one row per client-specific treatment episode and include variables such as:

- `pre`
- `post`
- `change`
- `slope`
- `n_sessions`

These episode-level summaries can be passed directly to `plot_episode_change()`:

```{r episode-change, warning = FALSE, fig.width=5, fig.height=4, fig.align="center"}
plot_episode_change(episode_summary)
```

The plot displays individual client change scores across successive treatment episodes and superimposes the mean change for each episode number with approximate 95% confidence intervals.

Positive values represent improvement when the episode summaries were created so that higher values consistently indicate favorable change.

For larger samples, individual trajectories can be hidden to emphasize the episode-level summaries:

```{r episode-change-summary, warning = FALSE, fig.width=5, fig.height=4, fig.align="center"}
plot_episode_change(episode_summary, show_individuals = FALSE)
```

An important distinction is that `plot_episode_curves()` and `plot_episode_change()` operate at different levels of the data:

``` text
session-level data
  |- plot_episode_curves()
  |- episode_slopes()
    |- plot_episode_change()
```

`plot_episode_curves()` displays the observed treatment trajectories within episodes, whereas `plot_episode_change()` summarizes how the amount of treatment change varies across episode number.

# Plot episode decline

Treatment change can also occur outside active treatment. Clients may leave treatment at one level of functioning and return later with some degree of deterioration.

`episode_breaks()` summarizes these transitions between successive episodes:

```{r, episode-breaks, warning = FALSE, fig.width=5, fig.height=4, fig.align="center"}
breaks <- episode_breaks(episode_data)
```

Each row of the resulting data represents the transition from one treatment episode to the next for a client and includes variables such as:

- `prior_episode`
- `next_episode`
- `prior_discharge`
- `next_intake`
- `days_between`
- `bad_enough_level`
- `loss_per_month`

These data can be visualized using `plot_breaks_loss()`:

```{r, breaks-loss, warning = FALSE, fig.width=7, fig.height=4, fig.align="center"}
plot_episode_loss(breaks)
```

The plot connects the discharge score from one episode with the intake score at the subsequent episode. This provides a visual representation of the amount of outcome loss occurring between episodes.

When time is displayed, horizontal spacing can reflect the relative amount of time between episodes:

```{r, breaks-loss-time, warning = FALSE, fig.width=7, fig.height=4, fig.align="center"}
plot_episode_loss(breaks, show_time = TRUE)
```

Equal spacing can instead be used when the primary interest is the magnitude of change:

```{r breaks-loss-equal, warning = FALSE, fig.width=7, fig.height=4, fig.align="center"}
plot_episode_loss(breaks, show_time = FALSE)
```

Because only clients with multiple episodes can contribute between-episode transitions, a separate multiple-episode cohort restriction is not required.

# Summary

The visualization functions in `leap` provide complementary views of repeated treatment episodes:

- Use `plot_lag_density()` to inspect the temporal spacing of sessions and candidate episode delimiters.
- Use `plot_episode_curves()` to examine observed session-level trajectories within treatment episodes.
- Use `plot_episode_change()` to compare episode-level treatment change across successive episodes.
- Use `plot_breaks_loss()` to examine outcome loss occurring between episodes.

Together, these plots can help researchers assess the structure of multi-episode treatment data, identify potentially important patterns, and guide subsequent statistical modeling.
